Experimental and numerical studies of the effects of resonant and random excitations of the proton beam in the LHC, with applications to the design of a pulsed hollow electron lens for active halo control

### 2018-03-14 resex.tex
Intermediate step in revision of 8th-turn results. Moved figures.

@@

\subsection{$8^{\mathrm{th}}$ turn pulsing\label{sec:simex8}}

Pulsing patterns which have little effect on the beam core are
particularly interesting for the hollow electron lens operation if
they in addition have a large effect on the halo. This is in general
feasible considering the highly nonlinear field generated at the
location of the halo particles by the HEL compared to no field or only
a very small field at the location of the beam core. With this
motivation in mind the \eighthtp\ pattern was studied, for which a
small effect in comparison to the 7th- and \tenthtp\ is expected (see
Fig.~\ref{fig:8thsim}).

\begin{figure*}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2uran1_2e-3_8th_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2uran1_2e-3_8th_3_5um_sigm.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2uran1_2e-3_8th_3_5um_emit1.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2uran1_2e-3_8th_3_5um_emit2.png}
	\end{minipage}	
	\caption{\label{fig:8thsim} Relative intensity (top left), bunch length (top right), and horizontal (bottom left) and vertical (bottom right) emittance obtained from simulations (distribution tracking) based on the 2017 injection optics with standard errors and $(Q_x,Q_y)=(62.27,60.295)$. The solid line indicates an excitation with only a random dipole noise component in H+V of 6~nrad, and the dotted and dashed line the results for \eighthtp\ plus a random dipole noise component in H+V of 6~nrad.}
\end{figure*}
\begin{figure*}[t]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerru_dp0_amp.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$8^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerrut8skhv_dp0_amp_annotate.png}
	\end{minipage}	
	\caption{\label{fig:8th2017fmaamp} FMA analysis in amplitude space without excitation (left) and for \eighthtp\ (right) based on the 2017 injection optics with no machine errors and a tune of (62.27,60.295). The excitation is 96~nrad in both planes. The $16Q_y$ and \mbox{$8Q_x-4Q_y$} are both excited.}
\end{figure*}
In simulations no visible effect is indeed observed for excitation amplitudes smaller than 96~nrad. In the horizontal and vertical plane a small emittance growth due to the excitation is visible, but without any clear dependence on the amplitude or plane. The FMA analysis in amplitude space reveals also the driven resonances (Fig.~\ref{fig:8th2017fmaamp}), which are the $16Q_y$ and \mbox{$8Q_x-4Q_y$} resonances and several even higher orders. These are in general very high order resonances whose effect is thus expected to be small.

\begin{figure}[h]
	\centering
	\includegraphics[width=1.0\linewidth]{2017_scale_amp_8hv_lblshort.png}
	\caption{\label{fig:8thexploss} Comparison of scaling of loss rates with excitation amplitude for \eighthtp\ in H+V measured during the 2017 experiment. The excitation amplitude error is 15\% indicated by the horizontal error bars, the error on the loss rate is the error on the mean obtained through the average over the 6~bunches with the same excitation amplitude.}
\end{figure}
The small effect on the beam expected from simulations could then be also confirmed experimentally during the 2017 experiment. For this purpose the excitation amplitude was increased to the maximum reachable amplitude of 96~nrad. As depicted in Fig.~\ref{fig:8thexploss}, the loss rate increased slightly with the excitation amplitude for an excitation in H+V. For an excitation only in V or only in H, no statistically relevant increase was observed (see Appendix~\ref{app:sec:8}, Fig.~\ref{fig:8thexploss2017}). In the horizontal plane, a change in beam distribution was measured with the BSRT profiles (Appendix~\ref{app:sec:8}, Fig.~\ref{fig:8thexpprof}) for an excitation only in~H. The distribution increased around $2.0~\sigma$ simultaneously with a depletion of the beam core. For an excitation in H+V one would expect to see a similar distribution change. However, this change is not visible as the emittance of the references bunches was much smaller, around $1.8~\mu$m normalized emittance, compared to around $2.6~\mu$m for the excited bunches \cite{resexmd2017}. This difference in emittance then resulted in a stronger emittance growth due to intra-beam scattering for the reference bunches in comparison to the excited bunches making the additional emittance growth and distribution changes due to the excitation undetectable. In the vertical plane, the beam distribution and emittance were not affected by the resonant excitation.

\subsection{random excitation\label{sec:simexran}}

@@

\begin{figure*}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerru_dp0_amp.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		uniform random H+V
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerruranadthv_1nrad_dp0_amp.png}
	\end{minipage}	
	\caption{\label{fig:ran2017fmaamp} FMA analysis in amplitude space without excitation (left) and for a uniform random excitation with 1~nrad amplitude (right) in H+V based on the 2017 injection optics with no machine errors and a tune of (62.27,60.295).}
\end{figure*}

@@

\begin{figure*}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_sigm.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_emit1.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_emit2.png}
	\end{minipage}
	\caption{\label{fig:ransim} Relative intensity (top left), bunch length (top right), and horizontal (bottom left) and vertical (bottom right) emittance obtained from distribution tracking based on the 2017 injection optics with standard errors and $(Q_x,Q_y)=(62.27,60.295)$. The solid line indicates the case with no excitation, and the dotted and dashed line the results for a uniform random excitation of 12~nrad and 24~nrad respectively.}
\end{figure*}
The distribution tracking shown in Fig.~\ref{fig:ransim} also reveals a quite symmetric evolution of the beam parameters for a random excitation:
	\item Emittance growth is only caused by an excitation in the same plane.
	\item The emittance growth is quantitatively comparable for an excitation in only the corresponding plane. Only a small increase or decrease is observed if the excitation is applied in both planes. The difference could be due to coupling.
	\item The random excitation leads to an increase of the emittance growth rate. Compared to the $k^{\mathrm{th}}$~turn pulsing there is explicitly no initial adjustment phase of the distribution.
	\item Negligible losses are observed independent of the plane of excitation.
\end{itemize}
\begin{figure*}[b]
	random excitation V\\
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2017_bunch_intensity_vran_no_damper_avg.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2017_emith_avg_rel_vran_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2017_emitv_avg_rel_vran_no_damper.png}
	\end{minipage}	
		random excitation H+V\\
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2017_bunch_intensity_hvran_no_damper_avg.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2017_emith_avg_rel_hvran_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2017_emitv_avg_rel_hvran_no_damper.png}
	\end{minipage}	
	\caption{\label{fig:ranexp} Relative intensity (left) and relative horizontal (center) and vertical (right) emittance for a random excitation only in V (top) and in H+V (bottom). Emittance growth is only caused in the plane of excitation.}
\end{figure*}

@@

\begin{figure*}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		reference bunch,\\ no excitation
		\includegraphics[width=1.0\linewidth]{profile_v_ranv_slot_1698.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		excited bunch,\\ random excitation V
		\includegraphics[width=1.0\linewidth]{profile_v_ranv_slot_1532.png}
	\end{minipage}
	\caption{\label{fig:ranexpprof}Vertical beam profiles measured with the Beam Synchrotron Radiation Telescope (BSRT) during the 2017 experiments. The profiles are taken at the end of the random excitation in V. For both bunches the transverse damper is not active. The distribution changes in the reference bunch (left), here slot 1698, are much less than for the bunch experiencing the maximum excitation (right), here slot 2696.}
\end{figure*}
The changes in emittance are also directly visible in the beam profiles measured with the BSRT. As an example Fig.~\ref{fig:ranexpprof} shows the profile for a random excitation only in V in comparison to a reference bunch. The distribution also stays Gaussian in case of a random excitation in contrast to the \tenthtp\ in which case it clearly assumes a non-Gaussian shape (see Fig.~\ref{fig:10thexpprof}).

\begin{figure*}[h]
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_scale_amp_ranh_lblshort.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_scale_amp_ranv_lblshort.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_scale_amp_ranhv_lblshort.png}
	\end{minipage}	
	\caption{\label{fig:ranexplossplane} Comparison of scaling of loss rates with excitation amplitude for a random excitation only in H (top), only in V (center) and in H+V (bottom) measured during the 2017 experiment. The excitation amplitude error is 15\% indicated by the horizontal error bars. The loss rate is considerably reduced for the bunches with transverse damper active and is approximately additive with the excitation plane, meaning that double the loss rate is observed for an excitation in H+V compared to only in H or only in V.}
\end{figure*}
The experimental results for a maximum excitation amplitude of $A_{\mathrm{max}}=1$~nrad and $A_{\mathrm{max}}=6$~nrad also confirm the in general small losses predicted in simulations. The losses then increase more dramatically around the time when the maximum excitation amplitude is increased to $A_{\mathrm{max}}=12$~nrad (see Fig.~\ref{fig:ranexpprof}). This is most likely the point when, due to the large emittance growth, the physical aperture starts cutting into the more central region of the distribution. Furthermore, the dependence of the loss rate on the excitation plane is additive, meaning that double the loss rate is observed for an excitation in H+V compared to only in H or only V. The loss rate is also considerably reduced for the bunches with transverse damper active independent of the plane the excitation is applied in (see Fig.~\ref{fig:ranexplossplane}). In comparison to the \seventhtp, the loss rate for a random excitation is smaller for a maximum excitation amplitude of $A_{\mathrm{max}}=6$~nrad, but larger for $A_{\mathrm{max}}=12$~nrad (see Fig.~\ref{fig:ranexploss}). This indicates an increase of the diffusion mostly in the beam tails for \seventhtp, while for a random excitation mostly the core is affected.
\begin{figure}[h]
	\centering
	\includegraphics[width=1.0\linewidth]{2017_scale_amp_7th_ranhv_lblshort.png}
	\caption{\label{fig:ranexploss}  Comparison of scaling of loss rates with excitation amplitude for a random excitation and \seventhtp\ in H+V measured during the 2017 experiment. The excitation amplitude error is 15\% indicated by the horizontal error bars.}
\end{figure}

>>>>

\newcommand{\fma}{frequency-map analysis}

@@

\subsection{Pulsing every 8th turn}
\label{sec:simex8}

\begin{figure*}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2uran1_2e-3_8th_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2uran1_2e-3_8th_3_5um_sigm.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2uran1_2e-3_8th_3_5um_emit1.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2uran1_2e-3_8th_3_5um_emit2.png}
	\end{minipage}	
	\caption{\label{fig:8thsim} Relative intensity (top left), bunch length (top right), and horizontal (bottom left) and vertical (bottom right) emittance obtained from simulations (distribution tracking) based on the 2017 injection optics with standard errors and $(Q_x,Q_y)=(62.27,60.295)$. The solid line indicates an excitation with only a random dipole noise component in H+V of 6~nrad, and the dotted and dashed line the results for \eighthtp\ plus a random dipole noise component in H+V of 6~nrad.}
      \end{figure*}

\begin{figure*}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerru_dp0_amp.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$8^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerrut8skhv_dp0_amp_annotate.png}
	\end{minipage}	
	\caption{\label{fig:8th2017fmaamp} FMA analysis in amplitude space without excitation (left) and for \eighthtp\ (right) based on the 2017 injection optics with no machine errors and a tune of (62.27,60.295). The excitation is 96~nrad in both planes. The $16Q_y$ and \mbox{$8Q_x-4Q_y$} are both excited.}
\end{figure*}

% \begin{figure}
% 	\centering
% 	\includegraphics[width=1.0\linewidth]{2017_scale_amp_8hv_lblshort.png}
% 	\caption{\label{fig:8thexploss} Comparison of scaling of loss rates with excitation amplitude for \eighthtp\ in H+V measured during the 2017 experiment. The excitation amplitude error is 15\% indicated by the horizontal error bars, the error on the loss rate is the error on the mean obtained through the average over the 6~bunches with the same excitation amplitude.}
% \end{figure}

\begin{figure*}
  \begin{tabular}{ccc}
  \includegraphics[width=0.32\textwidth]{2017_scale_amp_8h_lblshort.png} &
  \includegraphics[width=0.32\textwidth]{2017_scale_amp_8v_lblshort.png} &
  \includegraphics[width=0.32\textwidth]{2017_scale_amp_8hv_lblshort.png} \\
  \end{tabular}
  \caption{Comparison of scaling of loss rates with excitation
    amplitude for \eighthtp\ only in H (top), only in V (center) and
    in H+V (bottom) during the 2017 experiment. The excitation
    amplitude error is 15\% indicated by horizontal error bars. The
    vertical errorbars represent the pure statistical error based on
    the averaging over several bunches with the same excitation
    amplitude.}
  \label{fig:8thexploss2017}
\end{figure*}

% \begin{figure}
% 	\begin{minipage}[t]{1.0\linewidth}
% 		\centering
% 		\includegraphics[width=0.9\linewidth]{2017_scale_amp_8h_lblshort.png}
% 	\end{minipage}
% 	\begin{minipage}[t]{1.0\linewidth}
% 		\centering
% 		\includegraphics[width=0.9\linewidth]{2017_scale_amp_8v_lblshort.png}
% 	\end{minipage}
% 	\begin{minipage}[t]{1.0\linewidth}
% 		\centering
% 		\includegraphics[width=0.9\linewidth]{2017_scale_amp_8hv_lblshort.png}
% 	\end{minipage}
% 	\caption{\label{fig:8thexploss2017} Comparison of scaling of loss rates with excitation amplitude for \eighthtp\ only in H (top), only in V (center) and in H+V (bottom) during the 2017 experiment. The excitation amplitude error is 15\% indicated by horizontal error bars. The vertical errorbars represent the pure statistical error based on the averaging over several bunches with the same excitation amplitude.}
% \end{figure}

\begin{figure}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		reference bunch,\\ no excitation
		\includegraphics[width=1.0\linewidth]{profile_h_8thh_slot_804.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		excited bunch,\\ $8^{\mathrm{th}}$~turn H
		\includegraphics[width=1.0\linewidth]{profile_h_8thh_slot_632.png}
	\end{minipage}
	\caption{\label{fig:8thexpprof} Horizontal beam profiles measured with the Beam Synchrotron Radiation Telescope (BSRT) during the 2017 experiments. The profiles are taken at the end of the \eighthtp\ in H. For both bunches the transverse damper is not active. The distribution changes in the reference bunch (left), here slot 804, are much less than for the bunch experiencing the maximum excitation (right), here slot 632.}
\end{figure}

Excitation patterns that have little effect on the beam core and a
large effect on the halo are good candidates for extending the range
of hollow electron lens operation. In principle, this is possible
because of the highly nonlinear fields generated by the HEL at the
transverse locations of the beam halo and the small fields at the beam
core. For this reason, the \eighthtp\ pattern was chosen for the
experimental studies, as simulations indicated much smaller core
effects compared to 7th- and \tenthtp, as discussed in
Section~\ref{sec:pattern}.

Simulations of intensities, emittances, and bunch lengths for
different amplitudes of the 8th-turn excitation are shown in
Fig.~\ref{fig:8thsim}. No significant effects are predicted for
excitation amplitudes up to 192~nrad. Horizontal and vertical
emittances show small changes in growth rate, but without a clear
dependence on the amplitude or plane of the excitation.

The \fma\ in amplitude space is shown in
Fig.~\ref{fig:8th2017fmaamp}. It reveals that the driven resonances
are of high order, mainly $16 Q_y$ and \mbox{$8 Q_x - 4 Q_y$}, and
their effect is therefore expected to be small.

The small effect on the beam expected from simulations could then be
also confirmed experimentally during the 2017 experiment. For this
purpose the excitation amplitude was increased to the maximum
reachable amplitude of 96~nrad. As depicted in
Fig.~\ref{fig:8thexploss}, the loss rate increased slightly with the
excitation amplitude for an excitation in H+V. For an excitation only
in V or only in H, no statistically relevant increase was observed
(see Appendix~\ref{app:sec:8}, Fig.~\ref{fig:8thexploss2017}). In the
horizontal plane, a change in beam distribution was measured with the
BSRT profiles (Appendix~\ref{app:sec:8}, Fig.~\ref{fig:8thexpprof})
for an excitation only in~H. The distribution increased around
$2.0~\sigma$ simultaneously with a depletion of the beam core. For an
excitation in H+V one would expect to see a similar distribution
change. However, this change is not visible as the emittance of the
references bunches was much smaller, around $1.8~\mu$m normalized
emittance, compared to around $2.6~\mu$m for the excited bunches
\cite{resexmd2017}. This difference in emittance then resulted in a
stronger emittance growth due to intra-beam scattering for the
reference bunches in comparison to the excited bunches making the
additional emittance growth and distribution changes due to the
excitation undetectable. In the vertical plane, the beam distribution
and emittance were not affected by the resonant excitation.

% The small effect on the beam expected from simulations could then be
% also confirmed experimentally during the 2017 experiment. For this
% purpose the excitation amplitude was increased to the maximum
% reachable amplitude of 96~nrad. As depicted in
% Fig.~\ref{fig:8thexploss}, the loss rate increased slightly with the
% excitation amplitude for an excitation in H+V. For an excitation only
% in V or only in H, no statistically relevant increase was observed
% (see Appendix~\ref{app:sec:8}, Fig.~\ref{fig:8thexploss2017}). In the
% horizontal plane, a change in beam distribution was measured with the
% BSRT profiles (Appendix~\ref{app:sec:8}, Fig.~\ref{fig:8thexpprof})
% for an excitation only in~H. The distribution increased around
% $2.0~\sigma$ simultaneously with a depletion of the beam core. For an
% excitation in H+V one would expect to see a similar distribution
% change. However, this change is not visible as the emittance of the
% references bunches was much smaller, around $1.8~\mu$m normalized
% emittance, compared to around $2.6~\mu$m for the excited bunches
% \cite{resexmd2017}. This difference in emittance then resulted in a
% stronger emittance growth due to intra-beam scattering for the
% reference bunches in comparison to the excited bunches making the
% additional emittance growth and distribution changes due to the
% excitation undetectable. In the vertical plane, the beam distribution
% and emittance were not affected by the resonant excitation.

\subsection{Random excitation}
\label{sec:simexran}

\begin{figure*}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerru_dp0_amp.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		uniform random H+V
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerruranadthv_1nrad_dp0_amp.png}
	\end{minipage}	
	\caption{\label{fig:ran2017fmaamp} FMA analysis in amplitude space without excitation (left) and for a uniform random excitation with 1~nrad amplitude (right) in H+V based on the 2017 injection optics with no machine errors and a tune of (62.27,60.295).}
\end{figure*}

\begin{figure*}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_sigm.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_emit1.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_emit2.png}
	\end{minipage}
	\caption{\label{fig:ransim} Relative intensity (top left), bunch length (top right), and horizontal (bottom left) and vertical (bottom right) emittance obtained from distribution tracking based on the 2017 injection optics with standard errors and $(Q_x,Q_y)=(62.27,60.295)$. The solid line indicates the case with no excitation, and the dotted and dashed line the results for a uniform random excitation of 12~nrad and 24~nrad respectively.}
      \end{figure*}

\begin{figure*}
	random excitation V\\
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2017_bunch_intensity_vran_no_damper_avg.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2017_emith_avg_rel_vran_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2017_emitv_avg_rel_vran_no_damper.png}
	\end{minipage}	
		random excitation H+V\\
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2017_bunch_intensity_hvran_no_damper_avg.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2017_emith_avg_rel_hvran_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2017_emitv_avg_rel_hvran_no_damper.png}
	\end{minipage}	
	\caption{\label{fig:ranexp} Relative intensity (left) and relative horizontal (center) and vertical (right) emittance for a random excitation only in V (top) and in H+V (bottom). Emittance growth is only caused in the plane of excitation.}
\end{figure*}

\begin{figure*}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		reference bunch,\\ no excitation
		\includegraphics[width=1.0\linewidth]{profile_v_ranv_slot_1698.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		excited bunch,\\ random excitation V
		\includegraphics[width=1.0\linewidth]{profile_v_ranv_slot_1532.png}
	\end{minipage}
	\caption{\label{fig:ranexpprof}Vertical beam profiles measured with the Beam Synchrotron Radiation Telescope (BSRT) during the 2017 experiments. The profiles are taken at the end of the random excitation in V. For both bunches the transverse damper is not active. The distribution changes in the reference bunch (left), here slot 1698, are much less than for the bunch experiencing the maximum excitation (right), here slot 2696.}
      \end{figure*}

\begin{figure*}
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_scale_amp_ranh_lblshort.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_scale_amp_ranv_lblshort.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_scale_amp_ranhv_lblshort.png}
	\end{minipage}	
	\caption{\label{fig:ranexplossplane} Comparison of scaling of loss rates with excitation amplitude for a random excitation only in H (top), only in V (center) and in H+V (bottom) measured during the 2017 experiment. The excitation amplitude error is 15\% indicated by the horizontal error bars. The loss rate is considerably reduced for the bunches with transverse damper active and is approximately additive with the excitation plane, meaning that double the loss rate is observed for an excitation in H+V compared to only in H or only in V.}
      \end{figure*}

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{2017_scale_amp_7th_ranhv_lblshort.png}
	\caption{\label{fig:ranexploss}  Comparison of scaling of loss rates with excitation amplitude for a random excitation and \seventhtp\ in H+V measured during the 2017 experiment. The excitation amplitude error is 15\% indicated by the horizontal error bars.}
\end{figure}

@@

@@

The distribution tracking shown in Fig.~\ref{fig:ransim} also reveals a quite symmetric evolution of the beam parameters for a random excitation:
\begin{itemize}
	\item Emittance growth is only caused by an excitation in the same plane.
	\item The emittance growth is quantitatively comparable for an excitation in only the corresponding plane. Only a small increase or decrease is observed if the excitation is applied in both planes. The difference could be due to coupling.
	\item The random excitation leads to an increase of the emittance growth rate. Compared to the $k^{\mathrm{th}}$~turn pulsing there is explicitly no initial adjustment phase of the distribution.
	\item Negligible losses are observed independent of the plane of excitation.
        \end{itemize}

@@

The changes in emittance are also directly visible in the beam profiles measured with the BSRT. As an example Fig.~\ref{fig:ranexpprof} shows the profile for a random excitation only in V in comparison to a reference bunch. The distribution also stays Gaussian in case of a random excitation in contrast to the \tenthtp\ in which case it clearly assumes a non-Gaussian shape (see Fig.~\ref{fig:10thexpprof}).

The experimental results for a maximum excitation amplitude of
$A_{\mathrm{max}}=1$~nrad and $A_{\mathrm{max}}=6$~nrad also confirm
the in general small losses predicted in simulations. The losses then
increase more dramatically around the time when the maximum excitation
amplitude is increased to $A_{\mathrm{max}}=12$~nrad (see
Fig.~\ref{fig:ranexpprof}). This is most likely the point when, due to
the large emittance growth, the physical aperture starts cutting into
the more central region of the distribution. Furthermore, the
dependence of the loss rate on the excitation plane is additive,
meaning that double the loss rate is observed for an excitation in H+V
compared to only in H or only V. The loss rate is also considerably
reduced for the bunches with transverse damper active independent of
the plane the excitation is applied in (see
Fig.~\ref{fig:ranexplossplane}). In comparison to the \seventhtp, the
loss rate for a random excitation is smaller for a maximum excitation
amplitude of $A_{\mathrm{max}}=6$~nrad, but larger for
$A_{\mathrm{max}}=12$~nrad (see Fig.~\ref{fig:ranexploss}). This
indicates an increase of the diffusion mostly in the beam tails for
\seventhtp, while for a random excitation mostly the core is affected.

\subsection{Effect of the transverse damper}
\label{sec:damp}

\begin{figure*}
	\centering
	damper not active\\
	\begin{minipage}[t]{0.32\linewidth}
		\centering
	random excitation V, 2017\\
		\includegraphics[height=0.75\linewidth]{2017_emitv_avg_rel_vran_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		7th~turn V, 2017 exp.\\
		\includegraphics[height=0.75\linewidth]{2017_emitv_avg_rel_v7th_no_damper_no_text.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		$10^{\mathrm{th}}$~turn V, 2016 exp.\\
		\includegraphics[height=0.75\linewidth]{2016_emitv_avg_rel_v10th_no_damper_no_text.png}
	\end{minipage}	
	damper active\\
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		random excitation V, 2017 exp.\\
		\includegraphics[height=0.75\linewidth]{2017_emitv_avg_rel_vran_with_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		7th~turn V, 2017 exp.\\
		\includegraphics[height=0.75\linewidth]{2017_emitv_avg_rel_v7th_with_damper_no_text.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		$10^{\mathrm{th}}$~turn V, 2016 exp.\\
		\includegraphics[height=0.75\linewidth]{2016_emitv_avg_rel_v10th_with_damper_no_text.png}
	\end{minipage}	
	\caption{\label{fig:damp} Relative intensity for bunches with transverse damper not active (top) and transverse damper active (bottom) for a random excitation only in V (left) and for 7th- (center) and \tenthtp\ only in V (right).}
      \end{figure*}

For all $k^{\mathrm{th}}$~turn pulsing patterns tested experimentally,
the transverse damper does in general not change any of the
observables studied during the experiments, explicitly losses,
emittance and beam distribution suggesting that indeed it is not
damping any resonant excitation. On the other hand, the transverse
damper considerably decreases any changes of the above parameters in
case of a random excitation. As an example, the vertical emittance
growth for the 7th- and \tenthtp\ in V is compared with a random excitation in V in Fig.~\ref{fig:damp}. The same observation is also made in case of losses as shown in Fig.~\ref{fig:ranexploss} comparing the \seventhtp\ in H+V with a random excitation in H+V. For the \seventhtp\ the two curves for the transverse damper active and not active almost coincide while for the random excitation a significant difference is visible. Until present, it is unknown why the transverse damper (ADT) appears to be incapable of damping any resonant excitation as it fulfills all requirements to detect as well as damp the caused oscillation:
\begin{itemize}
	\item The closed orbit distortion caused by the resonant excitation during the experiments is large enough to be detectable with the ADT pickups.
	\item The ADT compares the position of each bunch with its position in the previous turn in order to detect an orbit distortion to be damped. Therefore an oscillation due to any $k^{\mathrm{th}}$~turn pulsing with $k>1$ would be noticed and damped.
	\item The ADT is capable of damping the oscillation of each bunch individually. The ADT should therefore be capable of applying the correct corrective kick for each group of bunches seeing the same excitation amplitude.
\end{itemize}
Further studies are therefore needed in order to obtain a deeper understanding of the interaction of the transverse damper with a resonant excitation.

\section{Summary}
\label{sec:sum}

This paper intends to give an overview of the effects of the hollow electron lens (HEL) on the proton beam core and the experiments conducted in this context at the LHC in 2016~\cite{resexmd2016} and 2017~\cite{resexmd2017}. For a perfect HEL without any imperfections no effect on the core is to be expected as the field at the center vanishes by construction. Only in the case of imperfections in the guidance and profile of the electron beam, a residual kick on the proton beam core can occur. In this case and assuming HL-LHC parameters (Tables~\ref{tab:hllhc_param}--\ref{tab:hel_param}), the contribution from the electron lens bends with approximately $0.5$~nrad to first order is negligible with respect to the kick expected from the central region of the HEL of approximatley $20$~nrad to first order caused by profile imperfections in the electron beam. In DC operation these kicks, even if nonlinear in reality, are negligible compared to the nonlinearities present in the LHC and HL-LHC and are thus of no concern. The picture changes if the HEL is operated in pulsed mode, in which case noise is introduced on the halo particles (intended) as well as on the beam core (unintended) in the presence of imperfections of the HEL entailing much more stringent tolerances.
